---
title: "LIFE748 - ML and data analysis"
author: "Ben LaCourse"
format: html
editor: visual
---

## Step 0: Load the required libraries

```{r warning=FALSE, message=FALSE}



library(DESeq2)          # Differential expression analysis
library(ggplot2)         # Plotting
library(EnhancedVolcano) # Volcano plots
library(pheatmap)        # Heatmaps
library(caret)           # Data partitioning and confusion matrices
library(MASS)            # LDA
library(e1071)           # SVM
library(dplyr)           # Data manipulation
library(tibble)          # Row names / tibble helpers
library(knitr)           # Tables
```

If these packages are not already installed, they can be installed from CRAN or Bioconductor. To maximise reproducibility, it is also useful to record the R and package versions used when rendering the final document.

```{r}
set.seed(123)
```

## Step 1: Load the count matrix

The first 25 samples are assumed to belong to Condition A and the next 25 to Condition B.

```{r}
counts <- read.csv("dataset_LIFE748_3.csv", row.names = 1, check.names = FALSE)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = factor(c(rep("CondA", 25), rep("CondB", 25)),
                     levels = c("CondA", "CondB"))
)

# Basic checks
cat("Number of genes:", nrow(counts), "\n")
cat("Number of samples:", ncol(counts), "\n")
table(colData$Condition)
```

## Step 2: Normalize and identify DEGs with DESeq2

```{r}
# Create DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = colData,
  design    = ~ Condition
)

# Run DESeq2 pipeline
dds <- DESeq(dds)

# Extract DE results (Condition B vs Condition A)
res <- results(dds, contrast = c("Condition", "CondB", "CondA"))

# Variance-stabilised expression for downstream ML
vsd <- vst(dds, blind = FALSE)
normalized_counts <- assay(vsd)

# Convert results to data frame
res_df <- as.data.frame(res) %>%
  rownames_to_column("Gene") %>%
  filter(!is.na(padj)) %>%
  arrange(padj)

# Add average expression by condition for interpretation
res_df <- res_df %>%
  mutate(
    mean_CondA = rowMeans(normalized_counts[Gene, colData$Condition == "CondA", drop = FALSE]),
    mean_CondB = rowMeans(normalized_counts[Gene, colData$Condition == "CondB", drop = FALSE]),
    direction = ifelse(log2FoldChange > 0, "Higher in CondB", "Higher in CondA")
  )

# Overview of DE results
summary(res)
```

## Step 3: Report DEGs (table and volcano plot)

Statistically significant DEGs are defined here as genes with adjusted *p* value \< 0.05 and absolute log2 fold change \> 1.

```{r}
# Stringent DEG definition for reporting
sig_deg <- res_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1) %>%
  arrange(padj)

cat("Number of significant DEGs (padj < 0.05 and |log2FC| > 1):", nrow(sig_deg), "\n")

# Table of top DEGs for the report
sig_deg_table <- sig_deg %>%
  select(Gene, log2FoldChange, lfcSE, stat, pvalue, padj, mean_CondA, mean_CondB, direction) %>%
  slice_head(n = 20)

kable(sig_deg_table, digits = 3,
      caption = "Top 20 significant differentially expressed genes ranked by adjusted p-value.")

# Save full DEG table for supplementary output
write.csv(sig_deg, "DEG_table_stringent.csv", row.names = FALSE)
```

Generate a volcano plot to visualise the results.

```{r fig.width=9, fig.height=7}
EnhancedVolcano(
  res_df,
  lab = res_df$Gene,
  x = "log2FoldChange",
  y = "padj",
  pCutoff = 0.05,
  FCcutoff = 1,
  pointSize = 2.0,
  labSize = 3.0,
  title = "Volcano plot: Condition B vs Condition A",
  subtitle = "DESeq2 differential expression results",
  caption = "Significant DEGs satisfy padj < 0.05 and |log2FC| > 1"
)
```

## Step 4: Isolate differentially expressed genes

```{r}
# Relaxed DEG threshold for clustering / classification
relaxed_deg <- res_df %>%
  filter(padj < 0.10, abs(log2FoldChange) > 0.5) %>%
  arrange(padj)

cat("Number of relaxed DEGs (padj < 0.10 and |log2FC| > 0.5):", nrow(relaxed_deg), "\n")

# Extract relaxed DEG expression matrix
deg_genes  <- relaxed_deg$Gene
deg_counts <- normalized_counts[deg_genes, , drop = FALSE]

cat("Dimensions of DEG-only expression matrix:", nrow(deg_counts), "genes x", ncol(deg_counts), "samples\n")

write.csv(deg_counts, "DEG_normalized_counts_relaxed.csv")
```

## Step 5: Unsupervised clustering

```{r}
# Transpose so rows = samples, columns = genes
deg_counts_t <- t(deg_counts)

# Standardise gene-wise for clustering and PCA
deg_counts_scaled <- scale(deg_counts_t)

# Remove any zero-variance columns that may break distance calculations
nzv <- apply(deg_counts_scaled, 2, sd, na.rm = TRUE) > 0
deg_counts_scaled <- deg_counts_scaled[, nzv, drop = FALSE]

cat("Number of relaxed DEGs retained after removing zero-variance features:", ncol(deg_counts_scaled), "\n")
```

### k-means clustering

```{r}
set.seed(123)
km <- kmeans(deg_counts_scaled, centers = 2, nstart = 50)

km_table <- table(Cluster = km$cluster, Condition = colData$Condition)
km_table

# Because k-means cluster labels are arbitrary, compute the best possible mapping accuracy
km_acc_1 <- sum(diag(km_table)) / sum(km_table)
km_acc_2 <- (km_table[1,2] + km_table[2,1]) / sum(km_table)
km_best_accuracy <- max(km_acc_1, km_acc_2)

cat("Best-mapped k-means clustering accuracy:", round(km_best_accuracy, 3), "\n")
```

### Hierarchical clustering

```{r fig.width=9, fig.height=6}
sample_dist <- dist(deg_counts_scaled, method = "euclidean")
hc <- hclust(sample_dist, method = "ward.D2")

plot(hc,
     labels = rownames(deg_counts_scaled),
     main = "Hierarchical clustering of samples",
     xlab = "",
     sub = "Ward.D2 linkage on scaled DEG expression",
     cex = 0.75)
rect.hclust(hc, k = 2, border = "red")
```

### Heatmap visualisation

```{r fig.width=10, fig.height=10}
annotation_col <- data.frame(
  Condition = colData$Condition,
  KmeansCluster = factor(km$cluster)
)
rownames(annotation_col) <- rownames(colData)

pheatmap(
  t(scale(t(deg_counts))),
  annotation_col = annotation_col,
  show_rownames = FALSE,
  show_colnames = FALSE,
  clustering_method = "ward.D2",
  main = "Heatmap of relaxed DEGs across samples"
)
```

### PCA for low-dimensional visualisation

```{r fig.width=8, fig.height=6}
pca <- prcomp(deg_counts_scaled, center = TRUE, scale. = FALSE)

pca_df <- data.frame(
  Sample = rownames(deg_counts_scaled),
  Condition = colData$Condition,
  Cluster = factor(km$cluster),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

pc_var <- summary(pca)$importance[2, 1:2] * 100

ggplot(pca_df, aes(PC1, PC2, color = Condition, shape = Cluster)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "PCA of samples using relaxed DEGs",
    x = paste0("PC1 (", round(pc_var[1], 1), "%)"),
    y = paste0("PC2 (", round(pc_var[2], 1), "%)")
  )
```

## Step 6: Supervised learning (logistic regression, LDA and SVM)

```{r}
# Prepare ML dataset: rows = samples, columns = genes
ml_data <- as.data.frame(t(deg_counts))
ml_data$Condition <- factor(colData$Condition, levels = c("CondB", "CondA"))

# Stratified split
set.seed(46)
train_index <- createDataPartition(ml_data$Condition, p = 0.70, list = FALSE)
train_data <- ml_data[train_index, ]
test_data  <- ml_data[-train_index, ]

cat("Training samples:", nrow(train_data), "\n")
cat("Testing samples:", nrow(test_data), "\n")

# Feature selection using training data only
x_train_all <- train_data[, setdiff(colnames(train_data), "Condition"), drop = FALSE]
feature_var <- apply(x_train_all, 2, var)
feature_var <- sort(feature_var, decreasing = TRUE)

# Keep a manageable number of highly variable genes
n_features <- min(20, length(feature_var))
selected_features <- names(feature_var)[1:n_features]

cat("Number of features used for supervised learning:", length(selected_features), "\n")
print(selected_features)

# Build scaled training and test matrices using identical scaling parameters
train_x <- scale(train_data[, selected_features, drop = FALSE])
train_center <- attr(train_x, "scaled:center")
train_scale  <- attr(train_x, "scaled:scale")

test_x <- scale(test_data[, selected_features, drop = FALSE],
                center = train_center,
                scale = train_scale)

train_model_data <- data.frame(as.data.frame(train_x), Condition = train_data$Condition)
test_model_data  <- data.frame(as.data.frame(test_x),  Condition = test_data$Condition)
```

### Train logistic regression

```{r}
glm_model <- glm(Condition ~ ., data = train_model_data, family = binomial)

glm_prob <- predict(glm_model, newdata = test_model_data, type = "response")
glm_class <- ifelse(glm_prob > 0.5, "CondA", "CondB")
glm_class <- factor(glm_class, levels = levels(test_model_data$Condition))
```

### Train LDA (Linear Discriminant Analysis)

```{r}
lda_model <- lda(Condition ~ ., data = train_model_data)
lda_pred  <- predict(lda_model, newdata = test_model_data)
lda_class <- lda_pred$class

plot(lda_model)
```

### Train SVM (Support Vector Machine)

```{r}
svm_model <- svm(
  Condition ~ ., 
  data = train_model_data,
  kernel = "linear",
  scale = FALSE,
  probability = TRUE
)

svm_class <- predict(svm_model, newdata = test_model_data)
plot(svm_class)
```

## Step 7: Compare classifier performance

Performance is evaluated on the unseen test set using confusion matrices and summary metrics.

```{r}
# Helper function to extract metrics from confusion matrices
extract_metrics <- function(cm, model_name) {
  data.frame(
    Model = model_name,
    Accuracy = unname(cm$overall["Accuracy"]),
    Kappa = unname(cm$overall["Kappa"]),
    Sensitivity = unname(cm$byClass["Sensitivity"]),
    Specificity = unname(cm$byClass["Specificity"]),
    BalancedAccuracy = unname(cm$byClass["Balanced Accuracy"]),
    Precision = unname(cm$byClass["Pos Pred Value"]),
    Recall = unname(cm$byClass["Sensitivity"]),
    F1 = unname(cm$byClass["F1"])
  )
}

true_labels <- test_model_data$Condition

glm_cm <- confusionMatrix(glm_class, true_labels, positive = "CondB")
lda_cm <- confusionMatrix(lda_class, true_labels, positive = "CondB")
svm_cm <- confusionMatrix(svm_class, true_labels, positive = "CondB")

glm_cm
lda_cm
svm_cm
```

```{r}
performance_summary <- bind_rows(
  extract_metrics(glm_cm, "Logistic regression"),
  extract_metrics(lda_cm, "LDA"),
  extract_metrics(svm_cm, "Linear SVM")
)

kable(performance_summary, digits = 3,
      caption = "Held-out test set performance for the three classifiers.")
```

## Step 8: Is this enough? Cross-validation and robustness analysis

```{r}
# Helper to create metrics from predictions
metric_from_preds <- function(pred, truth, model_name, positive = "CondB") {
  cm <- confusionMatrix(pred, truth, positive = positive)
  data.frame(
    Model = model_name,
    Accuracy = unname(cm$overall["Accuracy"]),
    Kappa = unname(cm$overall["Kappa"]),
    Sensitivity = unname(cm$byClass["Sensitivity"]),
    Specificity = unname(cm$byClass["Specificity"]),
    BalancedAccuracy = unname(cm$byClass["Balanced Accuracy"])
  )
}

set.seed(123)
folds <- createMultiFolds(ml_data$Condition, k = 5, times = 10)

cv_results <- list()
selected_gene_log <- list()

for (i in seq_along(folds)) {
  fold_train_index <- folds[[i]]
  fold_train <- ml_data[fold_train_index, , drop = FALSE]
  fold_test  <- ml_data[-fold_train_index, , drop = FALSE]
  
  # Training-only feature selection within each fold
  fold_x <- fold_train[, setdiff(colnames(fold_train), "Condition"), drop = FALSE]
  fold_var <- sort(apply(fold_x, 2, var), decreasing = TRUE)
  fold_n_features <- min(10, length(fold_var))
  fold_features <- names(fold_var)[1:fold_n_features]
  selected_gene_log[[i]] <- data.frame(Resample = names(folds)[i], Gene = fold_features)
  
  # Scale using fold training statistics only
  fold_train_x <- scale(fold_train[, fold_features, drop = FALSE])
  fold_center <- attr(fold_train_x, "scaled:center")
  fold_scale  <- attr(fold_train_x, "scaled:scale")
  
  fold_test_x <- scale(fold_test[, fold_features, drop = FALSE],
                       center = fold_center,
                       scale = fold_scale)
  
  fold_train_df <- data.frame(as.data.frame(fold_train_x), Condition = fold_train$Condition)
  fold_test_df  <- data.frame(as.data.frame(fold_test_x),  Condition = fold_test$Condition)
  
  # Logistic regression
  glm_fit <- glm(Condition ~ ., data = fold_train_df, family = binomial)
  glm_prob <- predict(glm_fit, newdata = fold_test_df, type = "response")
  glm_pred <- ifelse(glm_prob > 0.5, "CondA", "CondB")
  glm_pred <- factor(glm_pred, levels = levels(fold_test_df$Condition))
  
  # LDA
  lda_fit <- lda(Condition ~ ., data = fold_train_df)
  lda_pred <- predict(lda_fit, newdata = fold_test_df)$class
  
  # SVM
  svm_fit <- svm(Condition ~ ., data = fold_train_df, kernel = "linear", scale = FALSE)
  svm_pred <- predict(svm_fit, newdata = fold_test_df)
  
  # Store results
  cv_results[[length(cv_results) + 1]] <- metric_from_preds(glm_pred, fold_test_df$Condition, "Logistic regression")
  cv_results[[length(cv_results) + 1]] <- metric_from_preds(lda_pred, fold_test_df$Condition, "LDA")
  cv_results[[length(cv_results) + 1]] <- metric_from_preds(svm_pred, fold_test_df$Condition, "Linear SVM")
}

cv_results_df <- bind_rows(cv_results)
selected_gene_df <- bind_rows(selected_gene_log)
```

```{r}
cv_summary <- cv_results_df %>%
  group_by(Model) %>%
  summarise(
    Mean_Accuracy = mean(Accuracy, na.rm = TRUE),
    SD_Accuracy = sd(Accuracy, na.rm = TRUE),
    Mean_Sensitivity = mean(Sensitivity, na.rm = TRUE),
    Mean_Specificity = mean(Specificity, na.rm = TRUE),
    Mean_BalancedAccuracy = mean(BalancedAccuracy, na.rm = TRUE),
    Mean_Kappa = mean(Kappa, na.rm = TRUE),
    .groups = "drop"
  )

kable(cv_summary, digits = 3,
      caption = "Repeated 5-fold cross-validation performance summary.")
```

```{r fig.width=8, fig.height=6}
ggplot(cv_results_df, aes(x = Model, y = Accuracy)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Cross-validated accuracy across repeated resamples",
       x = "Classifier",
       y = "Accuracy")
```

## Biologically informative markers

To identify the most informative genes, differential expression evidence is combined with resampling-based feature selection frequency. Genes that are strongly differentially expressed and are repeatedly selected across cross-validation folds are more credible candidate markers than genes identified by only one method.

```{r}
# Frequency with which each gene was selected during cross-validation
selection_frequency <- selected_gene_df %>%
  count(Gene, sort = TRUE, name = "SelectionCount")

# Combine with DE statistics
marker_table <- selection_frequency %>%
  left_join(
    res_df %>% select(Gene, log2FoldChange, padj, mean_CondA, mean_CondB, direction),
    by = "Gene"
  ) %>%
  mutate(
    MeanDifference = mean_CondB - mean_CondA,
    AbsLog2FC = abs(log2FoldChange)
  ) %>%
  arrange(desc(SelectionCount), padj, desc(AbsLog2FC))

kable(head(marker_table, 20), digits = 3,
      caption = "Top candidate marker genes ranked by cross-validation selection frequency and supported by differential expression statistics.")

write.csv(marker_table, "candidate_marker_genes.csv", row.names = FALSE)
```

```{r}
# Optional: visualise the top marker genes
plot_markers <- head(marker_table$Gene, 15)
marker_plot_df <- res_df %>%
  filter(Gene %in% plot_markers) %>%
  mutate(Gene = reorder(Gene, log2FoldChange))

ggplot(marker_plot_df, aes(x = Gene, y = log2FoldChange, fill = direction)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top candidate marker genes",
       x = "Gene",
       y = "log2 fold change (CondB vs CondA)")
```

## Reproducibility outputs

```{r}
# Save key objects used in the analysis
write.csv(res_df, "DESeq2_all_results.csv", row.names = FALSE)
write.csv(performance_summary, "heldout_model_performance.csv", row.names = FALSE)
write.csv(cv_summary, "cross_validation_summary.csv", row.names = FALSE)
write.csv(selected_gene_df, "cv_selected_genes_by_resample.csv", row.names = FALSE)

sessionInfo()
```
